Reducing Acceptance Marks in Emerson-Lei Automata by QBF-solving

### 2022-04-26 paper/main.tex
fix typos

This section explains how we use the QBF formulas constructed in the
previous section to reduce the number of acceptance marks in TELA
automata. First we describe a \emph{single-level} reduction, which
uses only a single kind of QBF formulas. More precisely, we talk about
\emph{level~1}, \emph{level~2}, or \emph{level~3} reduction when
$\Phi^\mathit{opt}_{1,C,K}$, $\Phi^\mathit{opt}_{2,C,K}$, or
$\Phi^\mathit{opt}_{3,C,K}$ is used, respectively.

@@

The reduction procedure called $\mathit{SingleLevelReduction}$ is
given in Algorithm~\ref{alg:single}. Besides the reduction of
acceptance marks, the algorithm also reduces the number of clauses in
acceptance formula if the last argument $\mathit{reduceC}$ is set to
$\true$.
% In the first step, the procedure checks whether the input
% automaton represents the empty language and if this is the case, it
% returns the same automaton with acceptance conditon $\false$ and with
% all acceptance marks removed. Otherwise it
The first \textbf{while} loop gradually decreases the number of marks
until $K=1$ is reached or the QBF solver behind the function
$\mathit{satisfiable}(\Phi^\mathit{opt}_{j,C,K{-}1})$ fails to reduce
the number of marks, i.e., it claims unsatisfiability of the formula
or it runs out of resources. If the loop ends with $K=1$, we check
whether an acceptance condition without any mark (i.e., $\true$ or
$\false$) can be used. These checks are based on an inspection of the
automaton rather than on QBF solving. If some of the checks succeeds,
we return the correspodning automaton without any acceptance
mark. Otherwise, if $\mathit{reduceC}$ is set to $\true$ then the
procedure gradually reduces the number of clauses in the second
\textbf{while} loop. Note that the loop never checks for acceptance
condition with 0 clauses as it is equivalent to $\false$ and this case
was treated above. Finally, if the procedure succeeds to reduce the
number of marks or clauses, it constructs the modified
automaton. Otherwise, it returns the original automaton.

The algorithm can be reformulated to use an incremental approach
instead of building a new formula in each iteration of the
\textbf{while} loops. The incremental version of the first
\textbf{while} loop builds the formula
$\Phi=\Phi^\mathit{opt}_{j,C,K{-}1}$ only in the first iteration. In
each subsequent iteration, it extends this formula with a condition
saying that one more mark is not used in the automaton, i.e., the mark
is neither on edges, nor in the acceptance formula. For example, if we
want to say that the mark $k\in N_K$ is not used, we replace $\Phi$ by
\[
  \Phi~\wedge~\bigwedge_{t\in\delta}\neg n_{t,k}~\wedge\bigwedge_{c\in\{1,2,\ldots C\}}(\neg i_{c,k}\wedge\neg f_{c,k}).
\]
The second \textbf{while} can transformed to an incremental version
similarly. The incremental approach benefits from the fact that some
QBF solvers can decide an extended formula faster as they reuse the
information computed when solving the original formula.

@@

Figure~\ref{fig:example} shows a very simple automaton $\mA$ and the
three automata produced by calls of
$\mathit{SingleLevelReduction}(\mA,j,\true)$ for $j\in\{1,2,3\}$. The
figure clearly illustrates that the higher level of reduction we use,
the more acceptance marks can be reduced. On the other side, lower
levels are typically faster. The best results can be often achieved by
combining reductions of all levels. We call this aproach
\emph{multi-level} reduction. It is a straightforward sequential
application of the three levels, see Algorithm~\ref{alg:multi}.

@@

If some call of the function
$\mathit{satisfiable}(\Phi^\mathit{opt}_{j,C,K{-}1})$ in the first
\textbf{while} loop of Algorithm~\ref{alg:single} does not return
$\true$, then the name of the output automaton encodes the reason for
it. In case of a single level reduction, the name has the form
\texttt{L$j$\_$k$\_$X$}, where $j$ is the considered level, $k=K-1$ is
the number of acceptance mark considered by the formula, $X$ is either
\texttt{U} if the formula is unsatisfiable or \texttt{T} if the solver
ran out of resources. If $X$ is \texttt{T}, we know that additional
resources (typically longer timeout) may lead to further reductions.
If the multi-level reduction is used, the automaton name contains the
information from all levels. For example, the name `\texttt{L1\_5\_U
  L2\_3\_U L3\_1\_T}' means that level 1 reduced the number of marks
to 6 (reduction to 5 is impossible on this level), level 2 reduced it
to 4, and level 3 to 2 and the automaton may be potentially reduced
further.

@@

We have used the tool \texttt{genltl} of the Spot
library~\cite{duret.16.atva2} to generate LTL formulas from
literature. The parametrized formulas were generated for all
combinations of parameter values from 1 to 4. Altogether, we have
generated 348 unique LTL formulas. These have been translated to
deterministic TELA by two state-of-the-art translators from the Owl
library~\cite{kretinsky.18.atva} and to nondeterministic TELA by
another two state-of-the-art translators, namely \texttt{ltl2tgba}
from the Spot library~\cite{duret.16.atva2} and
\texttt{ltl3tela}~\cite{major.19.atva}. Each of the four automata sets
produced by individual tools contain less than 348 automata as not all
of the formulas are successfully translated due to timeout of 60
seconds. Further, we have removed automata with 0 or 1 acceptance
marks as there is little point reducing these. Table
\ref{tab:ltltranslators} shows the exact versions of used translators,
the homepage of the used tools and the number of automata produced by
the given tool from the LTL formulas after filtering.

@@

For each automata set (that is, the output of an individual LTL
translator), we have applied all single-level reductions and the
multi-level reduction, always with incremental approach and without
reduction of the number of clauses. The timeout for each QBF query was
set to 30 seconds.
%We have again filtered out the automata for which
%any of the four \telatko runs timed out after 120 seconds. We do this
%to make sure the numbers of acceptance marks can be meaningfully
%compared between different reduction levels of the same automata set.
%Additionally, we have filtered one automaton from the \texttt{ltl3tela} set
%that cannot be processed by \telatko due to an extensive use of system
%resources.

Table~\ref{tab:cumm-data} shows the cummulative numbers of marks in
the input automata sets and after each reduction, together with the
reduction ratio and total time spent by the considered reduction. The
columns \textit{solver response} and \textit{solver timeout} show the
number of automata for which the last query to QBF solver has been
successfully answered, or the solver timed out within a 30 seconds
limit to a single query, respectively. The timeout of the last QBF
query implies that the automaton could be potentially further reduced
by the reduction with a longer time limit.

@@

\begin{table}[h]
\caption{The cummulative number of acceptance marks before and after reduction
for various automata sets.
%\textit{X/Y solved aut.} means that out of
%\textit{Y} successfully translated LTL formulas, every run of \telatko ended
%within 120 seconds time limit on \textit{X} automata. We sum up the acceptance
%marks only on these automata.
The columns \textit{solver response} and
\textit{solver timeout} refer to last call to the QBF solver (that is, the
numbers of its successful responses within 30 seconds limit and timeouts).
}
\label{tab:cumm-data}
\centering
\input{table-data.tex}
\end{table}

@@

Figure~\ref{fig:quantile} presents the time spent by multi-level
reduction of automata in individual automata sets. The chart shows a
pleasing finding that for every set, most automata are reduced in
under a second and the high cummulative running times are caused by
few complicated automata.

@@

We have presented an application of QBF solving to reduction of the number of
acceptance marks in transition-based Emerson-Lei automata. The method
has been implemented and our experimental results show that it is possible
to use the QBF solvers to find equivalent acceptance conditions of TELA
with less acceptance marks and without changes to automata structure.

The reduction of acceptance marks is not the only application of the
presented approach: one could possibly look for, e.g., equivalent
parity automata or automata without $\Fin$-marks, depending on the
subsequent use of the automata. Even though the QBF queries can be
time-consuming, in practice it can be well used and one can find a
good trade-off between speed and efficiency by choosing precision of
the reduction and setting a reasonable timeout.

>>>>

This section explains how we use the QBF formulas constructed in the
previous section to reduce the number of acceptance marks in TELA
automata. First, we describe a \emph{single-level} reduction, which
uses only a single kind of QBF formulas. More precisely, we talk about
\emph{level~1}, \emph{level~2}, or \emph{level~3} reduction when
$\Phi^\mathit{opt}_{1,C,K}$, $\Phi^\mathit{opt}_{2,C,K}$, or
$\Phi^\mathit{opt}_{3,C,K}$ is used, respectively.

@@

The reduction procedure called $\mathit{SingleLevelReduction}$ is
given in Algorithm~\ref{alg:single}. Besides the reduction of
acceptance marks, the algorithm also reduces the number of clauses in the acceptance formula if the last argument $\mathit{reduceC}$ is set to
$\true$.
% In the first step, the procedure checks whether the input
% automaton represents the empty language and if this is the case, it
% returns the same automaton with acceptance conditon $\false$ and with
% all acceptance marks removed. Otherwise it
The first \textbf{while} loop gradually decreases the number of marks
until $K=1$ is reached or the QBF solver behind the function
$\mathit{satisfiable}(\Phi^\mathit{opt}_{j,C,K{-}1})$ fails to reduce
the number of marks, i.e., it claims unsatisfiability of the formula
or it runs out of resources. If the loop ends with $K=1$, we check
whether an acceptance condition without any mark (i.e., $\true$ or
$\false$) can be used. These checks are based on an inspection of the
automaton rather than on QBF solving. If some of the checks succeeds,
we return the corresponding automaton without any acceptance
mark. Otherwise, if $\mathit{reduceC}$ is set to $\true$ then the
procedure gradually reduces the number of clauses in the second
\textbf{while} loop. Note that the loop never checks for acceptance
condition with 0 clauses as it is equivalent to $\false$ and this case
was treated above. Finally, if the procedure succeeds to reduce the
number of marks or clauses, it constructs the modified
automaton. Otherwise, it returns the original automaton.

The algorithm can be reformulated to use an incremental approach
instead of building a new formula in each iteration of the
\textbf{while} loops. The incremental version of the first
\textbf{while} loop builds the formula
$\Phi=\Phi^\mathit{opt}_{j,C,K{-}1}$ only in the first iteration. In
each subsequent iteration, it extends this formula with a condition
saying that one more mark is not used in the automaton, i.e., the mark
is neither on edges, nor in the acceptance formula. For example, if we
want to say that the mark $k\in N_K$ is not used, we replace $\Phi$ by
\[
  \Phi~\wedge~\bigwedge_{t\in\delta}\neg n_{t,k}~\wedge\bigwedge_{c\in\{1,2,\ldots C\}}(\neg i_{c,k}\wedge\neg f_{c,k}).
\]
The second \textbf{while} can be transformed to an incremental version
similarly. The incremental approach benefits from the fact that some
QBF solvers can decide an extended formula faster as they reuse the
information computed when solving the original formula.

@@

Figure~\ref{fig:example} shows a very simple automaton $\mA$ and the
three automata produced by calls of
$\mathit{SingleLevelReduction}(\mA,j,\true)$ for $j\in\{1,2,3\}$. The
figure clearly illustrates that the higher level of reduction we use,
the more acceptance marks can be reduced. On the other side, lower
levels are typically faster. The best results can be often achieved by
combining reductions of all levels. We call this approach
\emph{multi-level} reduction. It is a straightforward sequential
application of the three levels, see Algorithm~\ref{alg:multi}.

@@

If some call of the function
$\mathit{satisfiable}(\Phi^\mathit{opt}_{j,C,K{-}1})$ in the first
\textbf{while} loop of Algorithm~\ref{alg:single} does not return
$\true$, then the name of the output automaton encodes the reason for
it. In the case of a single level reduction, the name has the form
\texttt{L$j$\_$k$\_$X$}, where $j$ is the considered level, $k=K-1$ is
the number of acceptance marks considered by the formula, $X$ is either
\texttt{U} if the formula is unsatisfiable or \texttt{T} if the solver
ran out of resources. If $X$ is \texttt{T}, we know that additional
resources (typically longer timeout) may lead to further reductions.
If the multi-level reduction is used, the automaton name contains the
information from all levels. For example, the name `\texttt{L1\_5\_U
  L2\_3\_U L3\_1\_T}' means that level 1 reduced the number of marks
to 6 (reduction to 5 is impossible on this level), level 2 reduced it
to 4, and level 3 to 2 and the automaton may be potentially reduced
further.

@@

We have used the tool \texttt{genltl} of the Spot
library~\cite{duret.16.atva2} to generate LTL formulas from
literature. The parametrized formulas were generated for all
combinations of parameter values from 1 to 4. Altogether, we have
generated 348 unique LTL formulas. These have been translated to
deterministic TELA by two state-of-the-art translators from the Owl
library~\cite{kretinsky.18.atva} and to nondeterministic TELA by
another two state-of-the-art translators, namely \texttt{ltl2tgba}
from the Spot library~\cite{duret.16.atva2} and
\texttt{ltl3tela}~\cite{major.19.atva}. Each of the four automata sets
produced by individual tools contains less than 348 automata as not all
of the formulas are successfully translated due to a timeout of 60
seconds. Further, we have removed automata with 0 or 1 acceptance
marks as there is little point in reducing these. Table
\ref{tab:ltltranslators} shows the exact versions of used translators,
the homepage of the used tools, and the number of automata produced by
the given tool from the LTL formulas after filtering.

@@

For each automata set (that is, the output of an individual LTL
translator), we have applied all single-level reductions and the
multi-level reduction, always with an incremental approach and without
reduction of the number of clauses. The timeout for each QBF query was
set to 30 seconds.
%We have again filtered out the automata for which
%any of the four \telatko runs timed out after 120 seconds. We do this
%to make sure the numbers of acceptance marks can be meaningfully
%compared between different reduction levels of the same automata set.
%Additionally, we have filtered one automaton from the \texttt{ltl3tela} set
%that cannot be processed by \telatko due to an extensive use of system
%resources.

Table~\ref{tab:cumm-data} shows the cumulative numbers of marks in
the input automata sets and after each reduction, together with the
reduction ratio and total time spent by the considered reduction. The
columns \textit{solver response} and \textit{solver timeout} show the
number of automata for which the last query to QBF solver has been
successfully answered, or the solver timed out within a 30 seconds
limit to a single query, respectively. The timeout of the last QBF
query implies that the automaton could be potentially further reduced
by the reduction with a longer time limit.

@@

\begin{table}[h]
\caption{The cumulative number of acceptance marks before and after reduction
for various automata sets.
%\textit{X/Y solved aut.} means that out of
%\textit{Y} successfully translated LTL formulas, every run of \telatko ended
%within 120 seconds time limit on \textit{X} automata. We sum up the acceptance
%marks only on these automata.
The columns \textit{solver response} and
\textit{solver timeout} refer to last call to the QBF solver (that is, the
numbers of its successful responses within 30 seconds limit and timeouts).
}
\label{tab:cumm-data}
\centering
\input{table-data.tex}
\end{table}

@@

Figure~\ref{fig:quantile} presents the time spent by multi-level
reduction of automata in individual automata sets. The chart shows a
pleasing finding that for every set, most automata are reduced in
under a second and the high cumulative running times are caused by
a few complicated automata.

@@

We have presented an application of QBF solving to reduction of the number of
acceptance marks in transition-based Emerson-Lei automata. The method
has been implemented and our experimental results show that it is possible
to use the QBF solvers to find equivalent acceptance conditions of TELA
with fewer acceptance marks and without changes to automata structure.

The reduction of acceptance marks is not the only application of the
presented approach: one could possibly look for, e.g., equivalent
parity automata or automata without $\Fin$-marks, depending on the
subsequent use of the automata. Even though the QBF queries can be
time-consuming, in practice it can be well used and one can find a
good trade-off between speed and efficiency by choosing the precision of
the reduction and setting a reasonable timeout.
